Constructing composite ordinal outcomes for clinical trials

### 2025-08-10 constructing_ordinal_outcomes.tex
Some restructuring and feedback

\title{Constructing ordinal outcomes for clinical trials}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

\begin{abstract}
  Ordinal outcomes are an increasingly popular choice for
  clinical trials.
  This paper is a guide to techniques for constructing ordinal
  outcomes out of pre-existing components, aimed at clinicians and their
  statistical collaborators.
  We discuss three broad mechanisms for combining existing binary,
  ordinal, or continuous
  outcomes into an ordinal composite---overriding (or
  ``sequential composition''), tie-breaking (or ``hierarchical
  composition''), and what we term ``parallel
  composition''---and illustrate each mechanism with a real-world example.
  We discuss how these mechanisms can improve the validity and
  efficiency of ordinal outcomes, but also why coarsening outcomes by
  combining levels might be preferred in practice.
  Our approach is closely related to the literature on prioritized
  outcomes (like win methods and
    desirability of outcome
  ranking methodologies), but we focus on the task of
  constructing ordinal
  outcomes and distinguish this from the related tasks of choosing
  effect measures and estimators and
  handling censoring.
\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Ordinal outcomes (or ``endpoints'') have become increasingly
prominent in clinical
resarch, particularly following the COVID-19 pandemic, when ordinal
scales developed by the WHO and NIAID were used in prominent trials
\citep{beigelRemdesivirTreatmentCovid192020,
marshallMinimalCommonOutcome2020}. Ordinal outcomes have also
been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020,
walkerCodesigningNovelOrdinal2025}.

Ordinal outcomes rank patient outcomes from better to worse, but
\emph{differences} in patient outcomes are
not treated as meaningful.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}.
\emph{Valid} outcomes reflect
patient benefit and harm; \emph{efficient} outcomes
are able to discriminate among better- and worse-off
patients without large sample sizes.

@@

  \item \emph{Jointly summarizing efficacy and safety} by incorporating
    safety events into a
    primary outcome.

@@

For the purposes of this
article, set aside ease of collection, low missingess, objective
ascertainability, etc.

@@

Another goal is to situate the literature around ``generalized
pairwise comparisons of prioritized outcomes''
\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
methods \citep{evansDesirabilityOutcomeRanking2015,
ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
endpoints.

@@

\section{Reasons to combine (or fine-grain) ordinal outcomes}

@@

\subsection{Augmenting a state or time-to-event outcome with
longitudinal information}

@@

\section{Ways to combine ordinal outcomes}

\subsection{Sequential,
hierarchical, \& parallel composition}

@@

\subsection{Pairwise comparison methods (win and DOOR) implicitly construct
ordinal outcomes}

@@

\section{Reasons to coarsen ordinal outcomes}

@@

One point of view is that one would ``ideally'' analyze ordinal outcomes by
reporting the associated expected utilities \todo{cite Berry
  talk; example of
utility-weighted Rankin scale} and that since existing ordinal
methods supply
such utilities implicitly why not just supply them explicitly. A
counterpoint
is that the methods do not, in fact, supply utilities implicitly.
Another is that ordinal outcomes let one ``hedge'' about precise
utillities and this is OK:
fully-specified utility function may require more
extensive development and validation or may vary widely across patients.

>>>>

\title{Constructing composite ordinal outcomes for clinical trials}
\author{Leon Di Stefano + Ordinal
Working Group}
\date{\today}

@@

\begin{abstract}

  \todo{Suggested framing of this based on 2025-08-07 meeting with
    Kate and Rob is: advice on constructing ordinal outcomes from
    pre-existing components. In particular, ``constructing'' consists
    of the steps before ``validating'' (using external data). The two
    main headings are proposed to be: ``combining existing outcomes''
  and ``how many categories?''}

  Ordinal outcomes are an increasingly popular choice for clinical trials. This
  paper is a guide to techniques for constructing ordinal outcomes out of
  pre-existing components, aimed at clinicians and their statistical
  collaborators. We discuss three broad mechanisms for combining existing
  ordinal outcomes---including binary, continuous, and time to event
  outcomes---into an ordinal composite: overriding (or ``sequential
  composition''), tie-breaking (or ``hierarchical composition''), and what we
  term ``parallel composition''. We illustrate these mechanisms with examples
  from the literature and discuss how constructing fine-grained composite
  outcomes can improve validity and efficiency---but also why coarser outcomes
  might sometimes be preferred in practice. Finally, we connect these general
  methods for combining ordinal outcomes with the ``prioritized outcomes''
  literature (including win-based and desirability of outcome ranking (DOOR)
  methodologies), clearly delineating the task of constructing composite
  ordinal outcomes from the related but distinct tasks of choosing effect
  measures, choosing estimators, and handling censoring.

\end{abstract}
\newpage
\tableofcontents
\newpage

@@

Ordinal outcomes (or endpoints) rank patients trajectories from better to
worse, but do not treat \emph{differences} in individual patient outcomes as
meaningful. Some ordinal outcomes have a long history in specific disease
areas, for example the modified Rankin scale in acute stroke trials
\citep{broderickEvolutionModifiedRankin2017}, and ordinal outcomes have become
increasingly prominent in clinical resarch following the COVID-19 pandemic when
disease-specific ordinal outcome scales were developed by the WHO and NIAID
\citep{beigelRemdesivirTreatmentCovid192020, marshallMinimalCommonOutcome2020}.
Ordinal outcomes have also been used in a number of adaptive platform trials
\citep{angusREMAPCAPRandomizedEmbedded2020, walkerCodesigningNovelOrdinal2025}.

Two desirable features of outcomes in general are \emph{validity} and
\emph{efficiency}. \emph{Valid} outcomes reflect patient benefit and harm;
\emph{efficient} outcomes are able to discriminate among better- and worse-off
patients with relatively small sample sizes. (Strictly speaking efficiency is a
  property not of outcomes themselves but of estimators of effect measures for
  summarizing differences in outcome distribution---more on this in
\ref{sec:prioritized_outcomes}.)

@@

  \item \emph{Jointly summarizing efficacy and safety} by incorporating
    safety events into a
    primary outcome; this might be particularly useful in pragmatic
    contexts or where side-effects of an intervention are difficult to identify.

@@

@@

Another goal is to situate the literature around ``generalized pairwise
comparisons of prioritized outcomes''
\citep{buyseGeneralizedPairwiseComparisons2022}--including win ratio
\citep{pocockWinRatioNew2012} and Desirability of Outcome Ranking (DOOR)
methods \citep{evansDesirabilityOutcomeRanking2015,
ongUnlockingDOORHow2023}--within the broader context of ordinal outcomes and
endpoints. These methods as typically presented implicitly bundle
together 3 distinct things:
(1) a choice of ordinal outcome or endpoint (composed through
tie-breaking in the terminology below); (2)
a choice of effect measure (a.k.a.~target parameter, estimand) based
on the ``independent pairs'' construction, and (3) a method for
handling censoring, particularly when the outcomes making up the
composite include time-to-event information.

@@

\subsection{Reasons to use ordinal outcomes}

@@

\subsubsection{Augmenting a state or time-to-event outcome with
longitudinal information}

@@

\subsection{Gaining statistical efficiency}

(See below.)

\section{Three ways to create composite ordinal outcomes}

\subsection{Reviewing the literature for existing/core outcomes}

\subsection{Overriding, tie-breaking, \& parallel composition}

@@

\section{How many levels?}

\subsection{Consider efficiency within levels of strong predictors}

\section{Hierarchical ordinal outcomes and pairwise comparison methods}

@@

@@

For the purposes of this article we have set aside other desiderata for
outcomes like ease of collection, low missingess, objective ascertainability,
etc.

One point of view is that one would ``ideally'' analyze ordinal outcomes by
reporting the associated expected utilities \todo{cite Berry talk; example of
utility-weighted Rankin scale} and that since existing ordinal methods supply
such utilities implicitly why not just supply them explicitly. A counterpoint
is that the methods do not, in fact, supply utilities implicitly. Another is
that ordinal outcomes let one ``hedge'' about precise utilities and this is
OK: fully-specified utility function may require more extensive development and
validation or may vary widely across patients.

\subsection{Censoring, consistent measurement, etc. are all
additional aspects of constructing the outcome that need to be considered}
